---
title: "grmbayes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grmbayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(grmbayes)
```

# Introduction

`grmbayes` provides a suite of functions to fit Bayesian geostatistical regression models, originally designed for use with air quality data. 

There are three primary functions in the package:
- `grm`: fits a Bayesian geostatistical regression model using MCMC
- `grm_cv`: fits a Bayesian geostatistical regression model using MCMC with cross-validation
- `grm_pred`: predicts values at new locations using a fitted Bayesian geostatistical regression model 

Examples of each of these functions are provided in this vignette using a dataset of $PM_{2.5}$ and Chemical Transport Model (CTM) readings collected in the Atlanta Metropolin Area from 2003 to 2005. Run `?cmaq_aqs_matched` for more info. 
A dataset of $PM_{2.5}$ and satellite-collected Aerosol Optical Depth (AOD) over the same area is also included in this package (`modis_aqs_matched`) but is not used in this vignette.


# grm

`grm()` fits a Bayesian geostatistical regression model using MCMC, detailed as follows:

\[
Y(s,t) = \alpha(s, t) + \beta(s, t) X(s, t) + \gamma L + \delta M + \epsilon(s, t) \\
\]
where
\[
        \alpha(s, t) = \alpha(s) + \alpha(t) \\
        \beta(s, t) = \beta(s) + \beta(t) 
\]

The model is specified in `grmbayes` using the following arguments:
- `Y`: a vector of response values (e.g., $PM_{2.5}$)
- `X`: a vector of primary covariate values (e.g., CTM)
- `L`: a data frame or matrix of secondary temporal covariate values (e.g., elevation, forest cover, etc.)
- `M`: a data frame or matrix of secondary spatio-temporal covariate values (e.g., temperature, wind speed, etc.)
- `coords`: a data frame or matrix of spatial coordinates, named `x` and `y`

`space.id`, `time.id`, and `spacetime.id` are also required to specify the model. 
These are vectors of unique identifiers for each location, time, and spatio-temporal combination, respectively. 
The unique identifiers must be integers starting at 1 and increasing by 1 for each additional location, time, or spatio-temporal combination


The following code provides a basic model fit:

```{r, eval = FALSE}
cmaq_fit <- grm(Y = cmaq_aqs_matched$pm25,
                X = cmaq_aqs_matched$ctm,
                L = cmaq_aqs_matched[, c("elevation", "forestcover",
                                 "hwy_length", "lim_hwy_length", 
                                 "local_rd_length", "point_emi_any")],
                M = cmaq_aqs_matched[, c("tmp", "wind")],
                coords = cmaq_aqs_matched[, c("x", "y")],
                space.id = cmaq_aqs_matched$space_id,
                time.id = cmaq_aqs_matched$time_id,
                spacetime.id = cmaq_aqs_matched$spacetime_id)
```

Output of `grm()` contains posterior draws of all fixed/random effects and model specification information. 

## Additive/Multiplicative Temporal/Spatial Effects

The user can specify whether the temporal and spatial random effects are additive or multiplicative using the `incl.add.temp.eff`, `incl.mult.temp.eff`, `incl.add.spat.eff` and `incl.mult.spat.eff` arguments.

For example, 

\[
Y(s,t) = \alpha(s, t) + \beta(s, t) X(s, t) + \gamma L + \delta M + \epsilon(s, t) \\
\]
where
\[
        \alpha(s, t) = \alpha(t) \\
        \beta(s, t) = \beta(s) 
\]

corresponds with 

```{r, eval = FALSE}
cmaq_fit <- grm(Y = cmaq_aqs_matched$pm25,
                X = cmaq_aqs_matched$ctm,
                L = cmaq_aqs_matched[, c("elevation", "forestcover",
                                 "hwy_length", "lim_hwy_length", 
                                 "local_rd_length", "point_emi_any")],
                M = cmaq_aqs_matched[, c("tmp", "wind")],
                coords = cmaq_aqs_matched[, c("x", "y")],
                space.id = cmaq_aqs_matched$space_id,
                time.id = cmaq_aqs_matched$time_id,
                spacetime.id = cmaq_aqs_matched$spacetime_id,
                incl.add.spat.eff = F,
                incl.add.temp.eff = T,
                incl.mult.spat.eff = T,
                incl.mult.temp.eff = F)
```

and 

\[
Y(s,t) = \alpha(s, t) + \beta(s, t) X(s, t) + \gamma L + \delta M + \epsilon(s, t) \\
\]
where
\[
        \alpha(s, t) = \alpha(t) \\
        \beta(s, t) = \beta(s) + \beta(t)
\]

corresponds with 

```{r, eval = FALSE}
cmaq_fit <- grm(Y = cmaq_aqs_matched$pm25,
                X = cmaq_aqs_matched$ctm,
                L = cmaq_aqs_matched[, c("elevation", "forestcover",
                                 "hwy_length", "lim_hwy_length", 
                                 "local_rd_length", "point_emi_any")],
                M = cmaq_aqs_matched[, c("tmp", "wind")],
                coords = cmaq_aqs_matched[, c("x", "y")],
                space.id = cmaq_aqs_matched$space_id,
                time.id = cmaq_aqs_matched$time_id,
                spacetime.id = cmaq_aqs_matched$spacetime_id,
                incl.add.spat.eff = F,
                incl.add.temp.eff = T,
                incl.mult.spat.eff = T,
                incl.mult.temp.eff = T)
```

## Spatial Processes

Nearest Neighbor Gaussian Processes (NNGP) can provide computational efficiency when there are many spatial locations. 
Choose either a Gaussian Process or Nearest Neighbor Gaussian Process for the spatial random effects using `nngp` argument. 
If `nngp = T`, one can also select number of neighbors (`num_neighbors`). 

```{r, eval = FALSE}
cmaq_fit <- grm(Y = cmaq_aqs_matched$pm25,
                X = cmaq_aqs_matched$ctm,
                L = cmaq_aqs_matched[, c("elevation", "forestcover",
                                 "hwy_length", "lim_hwy_length", 
                                 "local_rd_length", "point_emi_any")],
                M = cmaq_aqs_matched[, c("tmp", "wind")],
                coords = cmaq_aqs_matched[, c("x", "y")],
                space.id = cmaq_aqs_matched$space_id,
                time.id = cmaq_aqs_matched$time_id,
                spacetime.id = cmaq_aqs_matched$spacetime_id,
                incl.add.spat.eff = F,
                incl.add.temp.eff = T,
                incl.mult.spat.eff = T,
                incl.mult.temp.eff = F,
                nngp = T,
                num_neighbors = 10)
```




